---
title: "Data wrangling, regression analysis and model validation"
author: "Katariina Parnanen"
date: "November 7, 2018"
output:
  html_document: default
  github_document: default
---

##Data wrangling

###Read in data from local directory
```{r}
#Read in table
lrn2014<-read.table("/Users/kparnane/Documents/IODS_course/IODS-project/data/learning2014.txt", header=TRUE, sep="\t")

```
###Explore the dataset's dimension and structure
```{r}
#Explore the dimensions
dim(lrn2014)



#Explore the structure of the file
str(lrn2014)


```
The data has 166 observations and 7 variable columns.

The dataset has a response variable "points"" and explanatory variables "age", "attitude",  "deep", "surf" and "stra". "deep" means points for deep learning questions, "surf" points for surface learning questions, "stra" means strategic learning techniques. "points" means exam score. More info can be found [here](http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-meta.txt).

##Analysis

###Explore correlations and distributions of factors using "ggpairs"
```{r} 
library(GGally)
library(ggplot2)

#Draw plot
p <- ggpairs(lrn2014, mapping = aes(col = gender, alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))

p

#get summaries
summary(lrn2014$gender)
summary(lrn2014$age)
summary(lrn2014$attitude)
summary(lrn2014$deep)
summary(lrn2014$stra)
summary(lrn2014$surf)
```
There are more females than males. Other factors besides age seem to roughly follow a normal distribution. Mean age is 25. We are interested in the response variable points and how the explanatory variables corrrelate with it. The highest correlations are with "attitude", "stra" and there is a negative correlation with "surf".

###Linear models
Now we use the three variables with possible correlations with exam points identified in the data exploration.
```{r}
# fit a linear model
my_model <- lm(points ~ attitude+stra+surf, data = lrn2014)

# print out a summary of the model
summary(my_model)

```
Variable attitude is positively correlated with points with a p-value of 1.93e-08 and estimate 0.33952. The other variables are not significantly correlated with exam points so they can be excluded from the model.

###Drop the other variables and fit the model again
```{r}
# fit a linear model
my_model2 <- lm(points ~ attitude, data = lrn2014)

#Plot the regression line in scatter plot with exam points versus attitude
qplot(attitude, points, data = lrn2014) + geom_smooth(method = "lm")
```

###Summary of the model

```{r}
# print out a summary of the model
summary(my_model2)
```
There is a significant relationship with the variable attitude exam points. The attitude variable has a p-value of 1.95e-09, estimate of 0.35255 and standard error of  0.05674. The R-squared value is 0.1856 meaning that attitude variable explains roughly 19% of the variation in the model.

###Diagnostic plots
We will produce diagnostic plots to see if the model fits the assumptions of linear models.
The assumptions are that the size of the error is not dependent on the variable value and the errors are normally distributed and that the explanatory variables are not correlated with each other. The dependent variables should also be independent from each other.
```{r}
# Draw diagnostic plots using the plot() function. Choose the plots 1, 2 and 5
par(mfrow = c(2,2))
plot(my_model2, which=c(1, 2, 5))

```


There doesn't seem to be any patterns in the residuals vs fitted values plot so the size of the residuals are not dependent on the fitted values.

The residuals in the Q-Q plot follow the assumed regression line quite well so the assumption of normal distribution of errors is filled.

The residuals vs leverage plot doesn't reveal that there are any observations which have a unusually high effect on the model.

The model also has only one explanatory variable so explanatory variable autocorrelation is not a problem. The observations are also independent from each other.

Thus, we can conclude that the model fits the assumptions.

```{r}

```


