---
title: "Logistic regression"
author: "Katariina Parnanen"
date: "November 14, 2018"
output:
  html_document: default
  github_document: default
---
#Week 3: Logistic regression
This week we have learned about logistic regression, piping and model validation. 

##Reading in data

Read in data and explore.
```{r}
# Get access to useful libraries
library(dplyr);library(ggplot2);library(tidyr)

#Read in table from local folder
alc<-read.table(file = "data/alc.txt", sep="\t", header=TRUE)

#Explore the data using glimpse

glimpse(alc)


# Draw a bar plot of each variable
gather(alc) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()

```

The data has 382 observations and 35 variables from students whose alcohol consumption has been studied. The variables are either integers, numeric, factors or logicals. More information about the data can be found [here]https://archive.ics.uci.edu/ml/datasets/Student+Performance.

The data includes variables for sex, absences - absences from school, goout - going out and age. Usually there are differences between how much boys and girls drink due to social norms, absences from school also probably correlate with high use of alcohol, going out a lot also usually means higher alcohol consumption due to having more exposure to situations with alcohol and age might have an effect due to easier acccess to alcohol for older kids.

##Logistic regression

Logistic regression is one type of generalized linear models where the response variable can only have two values for example true or false. In this data set the response variable we are interested is high alochol use, which can have true or false values.

###Explore the relationship with high alcohol use and the chosen variables and the distribution of the variables

Explore how sex is related to high alcohol use. Our hypothesis is that males drink more than females.
```{r}
#Explore the distributions of females and males in the data
g0<-ggplot(data = alc, aes(x=sex))

g0 + geom_bar()

#Explore how high use is dependent on sex using barplots
g1<-ggplot(data = alc, aes(x = high_use, fill = sex))

g1 + geom_bar()+facet_wrap("sex")
```
There are about equal numbers of males and females in the data. There seems to be more males who have high use compared to females, which makes sense if you think about the social norms hypothesis

Explore how absences are related to alcohol use using boxplots. Our hypothesis is that kids who are absent more are more likely to also drink more.

```{r}

# Explore the distribution of absences
g6<-ggplot(alc, aes(x=absences, fill =sex))
g6+geom_bar() + ylab("absences") + ggtitle("Student absences by sex")


#Draw boxplots with high and low alcohol uses relation with absences
g3 <- ggplot(alc, aes(x=high_use, y=absences, col = sex))
g3 + geom_boxplot() +  ylab("absences") + ggtitle("Student absences by alcohol consumption and sex")

```

Absences are not normally distributed. There is a long tail and the most common value is 0. It seems that there are more absences in the kids with high alcohol use. This makes sense if you think that kids who drink more are also skipping more school. The patterns seem similar between females and males.

Explore how going out is related to alcohol use. Our hypothesis is that kids who are going out more are exposed to alcohol more and thus will also drink more.

```{r}
#Explore the distibution of kids going out
g4 <- ggplot(alc, aes(x=goout, fill=sex))
g4 + geom_bar()

#Explore the relationship with high alcohol use
g5 <- ggplot(alc, aes(x=high_use, y=goout, col = sex))
g5 + geom_boxplot()  + ggtitle("Going out by alcohol consumption and sex")
```

Going out seems to be quite normally distributed between the values 1 and 5. However, there aren't very many kids who don't go out almost at all compared to the ones who go out very often. There is approximately equal numbers of both sexes in each going out class.

There definitely seems to be a relationship between going out and alcohol use. However, there might be some differences between going out often beign correlated with drinking more since in females there is overlap between the ones who go out often but don't have high alcohol use with those who go out often and have high alcohol use.

Explore the relationship with age and alcohol use. Our hypothesis is that it is easier for older kids to get alcohol than for younger kids.

```{r}

# Explore the distribution of age
g6<-ggplot(alc, aes(x=age, fill =sex))
g6+geom_bar()

# Explore the relationship of age  with alcohol use

g7 <- ggplot(alc, aes(y=age, x=high_use, col = sex))
g7 + geom_boxplot() + ylab("age") + ggtitle("Student age by alcohol consumption and sex")
```
Age is not normally distributed and there is quite a long tail for higher numbers after 19. Ages range from 15 to 22 with 16 being the median. For the values 18 and up, there is an approximately equal amount of females and males for each age.

There doesn't seem to be a very obvious link between alcohol use and age that could be seen from the boxplots since the high use boxplots overlap with the low use bowplots. It might be that it is not very difficult to get alchol in the country of the kids even when you are younger than 18.


###Build model with the variables that seemed interesting

Next we will build a GLM using binomial family with the interesting explanatory variables.
```{r}
# Use generalized linear models to predict high use based on selected variables

m<-glm(high_use~age+sex+absences+goout, data=alc, family = "binomial")

#Print out summary and coefficients
summary(m)

#From the summary we can see that sex, absences and going out are significantly correlated with high alcohol use

coef(m)

#From the coefficients we can calculate the odds ratio by taking the exponent function of the estimate


# compute odds ratios (OR)
OR <- coef(m) %>% exp

# compute confidence intervals (CI)
CI<- confint(m) %>% exp

# Join the two together

cbind(OR, CI)


```
* From the odds ratio we can see that for every year the is approximately 9% increase in the chances of having high alcohol use, but the confidence interval includes one, which means that age is not significant.

* We can also see that sex male means a 2.6 times more likely chance of having high alcohol use. Female is the intercept factor. The CI does not include 1 so this going out is also significant.

* One more absence from school causes an approximately 9% bigger chance of having high use and this is significant also the CI doesn't include 1.

* Going out from a scale to 1 to 5 is significantly correlated with high alcohol use as the CI doesn't include 1. An increasement of 1 causes an approximately 2.6 more likely chance of having high use.

##Cross validation

After obtaining the model, we will use cross validation with a training set to estimate the error rate of the model. We will use the model without age since age was not significant.

```{r}

# Fit the model without age
m<-glm(high_use~sex+absences+goout, data=alc, family = "binomial")


# Predict the probability of high alcohol use
probabilities <- predict(m, type = "response")

# Add the predicted probabilities to the alc table
alc <- mutate(alc, probability = probabilities)

# Use the probabilities to make a prediction of high alcohol use as probabilitites instead of odds
alc <- mutate(alc, prediction = probabilities>0.5)

# See the last ten original classes, predicted probabilities, and class predictions
select(alc, failures, absences, sex, high_use, probability, prediction) %>% tail(10)

# From here you can see than when prob is >0.5 the prediction is "TRUE"

# Cross tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table %>% addmargins()



# From here we can see that out of 65/318 predictions are wrong for predicted low alcohol use and 15/64 are wrong for the prediction of high alcohol use.

# Define a loss function to get the mean prediction error
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# Call loss_func to compute the average number of wrong predictions in the (training) data
loss_func(class = alc$high_use, prob = alc$probability)


# Do K-fold cross-validation using the loss_func defined previously
library(boot)
cv <- cv.glm(data = alc, cost = loss_func, glmfit = m, K = 10)

# Average number of wrong predictions in the cross validation
cv$delta[1]



```

There is approximately 22 % error in the prediction of high alcohol use using the model wiht going out, absences and age.


